### 2025-05-21 docs/article/Sections/benchmark.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

The benchmark is designed with a modular architecture, separating the
\textit{view} and \textit{model} layers from the \textit{controller} layer
\cite{Bucanek2009}, which allows for easy extension and integration of new
template engines and frameworks. It also includes a set of tests to ensure the
correctness of implementations and to validate the \textit{HTML} output.

The benchmark includes two different data models, defined as \texttt{Presentation} and 
\texttt{Stock} shown in \autoref{lst:data-models}.

\lstdefinestyle{listingstyle}{
  backgroundcolor=\color{lightgray!20},
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{orange},
}

\lstset{style=listingstyle}

\begin{lstlisting}[language=Kotlin, caption={Data Models}, label={lst:data-models}]
data class Presentation(
    val id: Long,
    val title: String, 
    val speakerName: String,
    val summary: String
)

data class Stock(
    val name: String,
    val name2: String,
    val url: String,
    val symbol: String,
    val price: Double, 
    val change: Double, 
    val ratio: Double
)
\end{lstlisting}

@@

The Spring WebFlux core implementation uses Project Reactor to support a reactive
programming model: each method returns a \texttt{Flux<String>} as the response
body, which acts as a publisher that progressively streams the HTML content to
the client. It also includes methods using Kotlin coroutines and other
asynchronous mechanisms supported by template engines, such as the
\texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
using \textit{continuations}. 
The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we utilize
the \texttt{StreamingResponseBody} interface, which allows the application to
write directly to the response \texttt{OutputStream} without blocking the servlet
container thread. According to the Spring documentation, this class is \textit{a
controller method return value type for asynchronous request processing where
the application can write directly to the response OutputStream without holding
up the Servlet container thread} \cite{Spring-StreamingResponseBody}.

>>>>

The benchmark is designed with a modular architecture, separating the
\textit{view} and \textit{model} layers from the \textit{controller} layer
\cite{Bucanek2009}, which allows for easy extension and integration of new
template engines and frameworks. It also includes a set of tests to ensure the
correctness of implementations and to validate the \textit{HTML} output. It
includes two different data models, defined as \texttt{Presentation} and
\texttt{Stock}, shown in \autoref{lst:stock-class} and
\autoref{lst:presentation-class}, respectively. The \texttt{Presentation} class
represents a presentation with a title, speaker name, and summary, while the
\texttt{Stock} class represents a stock with a name, URL, symbol, price,
change, and ratio.

\vspace{0.5cm}

\lstset{style=listingstyle}
\begin{minipage}{0.5\textwidth}
  \begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={\texttt{Stock} class},
  label={lst:stock-class},
]
data class Stock(
    val name: String,
    val name2: String,
    val url: String,
    val symbol: String,
    val price: Double, 
    val change: Double, 
    val ratio: Double
)
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{0.4\textwidth}
  \begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={\texttt{Presentation} class},
    label={lst:presentation-class},
]
data class Presentation(
    val id: Long,
    val title: String, 
    val speakerName: String,
    val summary: String
)
\end{lstlisting}
\end{minipage}

@@

The Spring WebFlux core implementation uses Project Reactor to support a
reactive programming model: each method returns a \texttt{Flux<String>} as the
response body, which acts as a publisher that progressively streams the HTML
content to the client. The implementation includes three main approaches to
PSSR:
\begin{itemize}
  \item \textbf{Reactive}: The template engine is used in a reactive context,
        where the HTML content rendered using the reactive programming model.
        An example of this approach is the Thymeleaf template engine
        when using the \texttt{ReactiveDataDriverContextVariable} in conjunction with a
        non-blocking Spring \texttt{ViewResolver}.
  \item \textbf{Virtual}: The template engine is used in a non-blocking
        context, where the HTML content is rendered within the context of Virtual
        Threads. This method is used for the template engines that do not traditionally
        support non-blocking I/O, be it either because they use external DSLs and in
        consequence only support the blocking \texttt{Iterable} interface, or
        because they do not support the asynchronous rendering of HTML content.
  \item \textbf{Suspendable}: The template engine is used in a suspendable
        context, where the HTML content is rendered within the context of a suspending function
        An example of this approach is the HtmlFlow template engine, which
        supports suspendable templates with the use of the \texttt{Flow} class from the
        Kotlin standard library.
\end{itemize}
% It also includes methods using Kotlin coroutines and other
% asynchronous mechanisms supported by template engines, such as the
% \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
% using \textit{continuations}. 
The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we
utilize the \texttt{StreamingResponseBody} interface, which allows the
application to write directly to the response \texttt{OutputStream} without
blocking the servlet container thread. According to the Spring documentation,
this class is \textit{a controller method return value type for asynchronous
  request processing where the application can write directly to the response
  OutputStream without holding up the Servlet container thread}
\cite{Spring-StreamingResponseBody}.

### 2025-05-21 docs/article/Sections/introduction.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

% Non-blocking Progressive Server-Side Rendering (PSSR) is a technique that
% combines some of the advantages of both server-side and client-side rendering
% to improve web application performance and user experience. By leveraging
% non-blocking I/O operations, PSSR allows for efficient handling of multiple
% concurrent requests while progressively streaming HTML content to the client.
% 
% This project consists of a benchmark that evaluates the performance of
% different PSSR strategies on the Java Virtual Machine (JVM). The benchmark
% focuses on evaluating the performance of non-blocking PSSR implementations
% using approaches such as Kotlin coroutines, reactive-style programming, and
% Java virtual threads. Its main objective is to evaluate virtual threads as an
% alternative that ensures a non-blocking approach to PSSR while offering a more
% straightforward programming model than traditional asynchronous programming
% techniques.
% 
% This project aims to extend the previous work done in
% \textit{spring-webflux-comparing-template-engines}
% \citeauthor{spring-webflux-comparing-template-engines}, which compares the
% performance of various template engines in a Spring WebFlux application using
% different approaches to PSSR\@. It adds support for Java virtual threads and
% evaluates non-blocking PSSR implementations in other frameworks compared to
% Spring WebFlux.
% 
% Currently, this project features a working benchmark in Spring WebFlux, Spring
% MVC, and Quarkus, comparing the performance of eight different template
% engines: Rocker, JStachio, Pebble, Freemarker, Trimou, Thymeleaf, HtmlFlow, and
% KotlinX. It includes Java Microbenchmark Harness (JMH) benchmarks for template
% engine performance testing, as well as Apache Bench and Apache JMeter scripts
% for evaluating throughput and scalability. In the future, the project could be
% extended to include other frameworks such as Vert.x, if time allows. However,
% the next objective is the completion of the written report and article
% submission. The written report will include a more detailed analysis of the
% benchmark results, along with a contextualization of the state of the art in
% non-blocking PSSR and asynchronous programming.
% 
% This report will focus on detailing relevant aspects of the benchmark
% implementation and methodology, and will also provide a brief overview of the
% results obtained from the JMeter and Apache Bench tests, which evaluate the
% throughput and scalability of the different implementations.

>>>>

% Falta adicionar referências

Modern web applications rely on different rendering strategies to optimize
performance, user experience, and \textit{scalability}. The two most dominant
approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
    Rendering (CSR)}.

SSR generates HTML content on the server before sending it to the client,
resulting in faster initial page loads and better Search Engine Optimization
(SEO). However, SSR can increase server load and reduce
\textit{throughput}\footnote{The number of requests the server can handle per
    second (RPS)} since each request requires additional processing before
responding.

In contrast, CSR shifts the rendering workload to the browser. The server
initially sends a minimal HTML document with JavaScript, which dynamically
loads the page content. While CSR reduces the server’s burden, it can lead to
slower initial load times, as users must wait for JavaScript execution before
meaningful content appears.

\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
by streaming HTML content progressively. This technique enhances user-perceived performance by
allowing immediate rendering while data loads asynchronously. In this respect,
PSSR is similar to CSR in that the server initially sends a minimal HTML
document to the client and subsequently streams additional HTML fragments as data becomes available.
However, unlike CSR, PSSR retains all rendering responsibilities on the server side,
thereby reducing the load on the client. As a result, the client does
not need to execute JavaScript or make additional requests to retrieve page content.
To enable non-blocking PSSR, template engines require support for asynchronous
data models and non-blocking I/O operations. While some template engines, such
as HtmlFlow, and Thymeleaf, have support for these features, traditional
template engines that use \textit{external} domain-specific languages (DSLs)
often rely on blocking interfaces, such as \texttt{Iterable}, to define the
data model. This reliance on blocking interfaces leads to the system thread
being blocked until the entire HTML content is rendered, which can
significantly impact performance and scalability.

Efficient thread utilization is crucial for low-thread servers, necessitating
asynchronous, non-blocking request handling. In these architectures,
asynchronous I/O operations (e.g., database queries, and API calls) allow
servers to manage multiple concurrent requests efficiently by avoiding blocking
threads while waiting for I/O operations to complete. In these architectures,
where idle time during I/O operations is reduced, non-blocking PSSR enhances a
server's capacity to handle high request volumes with limited threads and
immediately send partial responses to clients.

With the introduction of \textit{Virtual Threads} in Java 21, it is now possible to execute blocking
I/O operations in a non-blocking manner, allowing for the use of these 
traditional template engines in non-blocking contexts without the need for
complex asynchronous programming models. This advancement enables the 
implementation of PSSR using familiar HTML syntax, simplifying template
development and adoption.

This work will focus on exploring existing approaches to non-blocking PSSR,
which include the use of \textit{reactive programming} and \textit{coroutines}
to implement non-blocking I/O operations. As an alternative to these
approaches, we will investigate the viability of \textit{Virtual Threads} in Java 21 
for implementing non-blocking PSSR\@. Section 2 provides an overview
of the state-of-the-art in non-blocking PSSR and HTML template engines. Section
3 provides a more detailed description of the issues associated with
traditional template engines for non-blocking PSSR, and how this work aims to
address them. Section 4 describes the benchmark methodology and implementation.
Finally, Sections 5 and 6 present the results and conclusions of this work,
respectively.

### 2025-05-21 docs/article/Sections/results.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

@@

The results in Figure~\ref{fig:presentations-webflux-jmeter} depict the
throughput (number of requests per second) for each template engine, with
concurrent users ranging from 1 to 128, from left to right. The benchmarks
include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
using Virtual Threads with the Iterable interface (Jstachio-Virtual), and Thymeleaf using the reactive View
Resolver driver (Thymeleaf-Rx). \textit{Sync} and \textit{Virtual} represent
the average throughput of the blocking approaches (i.e., KotlinX, Rocker,
Jstachio, Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the
context of a separate coroutine dispatcher or Virtual Threads, respectively.

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{Presentation Benchmark Results in Spring WebFlux with JMeter}\label{fig:presentations-webflux-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{Presentation Benchmark Results in Spring MVC with JMeter}\label{fig:presentations-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{Presentation Benchmark Results in Quarkus with JMeter}\label{fig:presentations-quarkus-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{Stocks Benchmark Results in Spring WebFlux with JMeter}\label{fig:stocks-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but replace the data model
with the more complex Stock class, including 20 instances. Despite the
increased complexity and number of instances, the scalability of the engines
remains largely unaffected. However, throughput is reduced by approximately 50
percent across all engines, with HtmlFlow achieving 6,000 requests per second.
The Thymeleaf implementation using the reactive View Resolver driver reaches
5,000 requests per second, suggesting that the performance drop compared to the
Presentation benchmark is not substantial.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{Stocks Benchmark Results in Spring MVC with JMeter}\label{fig:stocks-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{Stocks Benchmark Results in Quarkus with JMeter}\label{fig:stocks-quarkus-jmeter}
\end{figure}

@@

Additionally, the results show that approaches using Virtual Threads are able
to scale as effectively as those using reactive programming or Kotlin
coroutines, with the advantage of being easier to implement and understand for 
developers.

>>>>

Both the Quarkus and Spring MVC implementations were configured with an output
buffer size of 8KB, despite the Quarkus implementation not enabling PSSR at
this size. This configuration was chosen to ensure consistency across Spring
MVC and Quarkus, as the Spring MVC implementation does not support PSSR for the
tested templates.

@@

The results in Figure~\ref{fig:presentations-webflux-jmeter} depict the
throughput (number of requests per second) for each template engine, with
concurrent users ranging from 1 to 128, from left to right. The benchmarks
include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
using Virtual Threads with the \texttt{Iterable} interface (Jstachio-Virtual),
and Thymeleaf using the reactive View Resolver driver (Thymeleaf-Rx).
\textit{Sync} and \textit{Virtual} represent the average throughput of the
blocking approaches (i.e., KotlinX, Rocker, Jstachio, Pebble, Freemarker,
Trimou, HtmlFlow, and Thymeleaf) when run in the context of a separate
coroutine dispatcher or Virtual Threads, respectively.

@@

\begin{figure}[h]
     \isPreprints{\centering}{}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{Presentation Benchmark Results in Spring WebFlux with JMeter}\label{fig:presentations-webflux-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \isPreprints{\centering}{}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{Presentation Benchmark Results in Spring MVC with JMeter}\label{fig:presentations-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \isPreprints{\centering}{}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{Presentation Benchmark Results in Quarkus with JMeter}\label{fig:presentations-quarkus-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \isPreprints{\centering}{}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{Stocks Benchmark Results in Spring WebFlux with JMeter}\label{fig:stocks-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but replace the data model
with the more complex Stock class, including 20 instances. Despite the
increased complexity and number of instances, the scalability of the engines
remains largely unaffected. However, throughput is reduced by approximately 50
percent across all engines, with HtmlFlow achieving 6,000 requests per second.
Since the \texttt{Stock} class\ contains more data properties than the
\texttt{Presentation} class, the data binding to the template engine adds
overhead, which explains the reduced performance. The Thymeleaf implementation
using the reactive View Resolver driver reaches 5,000 requests per second,
suggesting that the performance drop compared to the Presentation benchmark is
not substantial.

\begin{figure}[h]
     \isPreprints{\centering}{}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{Stocks Benchmark Results in Spring MVC with JMeter}\label{fig:stocks-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \isPreprints{\centering}{}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{Stocks Benchmark Results in Quarkus with JMeter}\label{fig:stocks-quarkus-jmeter}
\end{figure}

@@

Additionally, the results show that approaches using Virtual Threads are able
to scale as effectively as those using reactive programming or Kotlin
coroutines, allowing \textit{external} DSLs to be used in a non-blocking context, 
effectively enabling PSSR\@.

### 2025-05-21 docs/project-proposal/ppG48.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

SSR generates HTML content on the server before sending it to the client, resulting in faster initial page loads and better Search Engine Optimization (SEO) \cite{prismic}. However, SSR can increase server load and reduce \textit{throughput}\footnote{The number of requests the server can handle per second (RPS)} since each request requires additional processing before responding.

In contrast, CSR shifts the rendering workload to the browser. The server initially sends a minimal HTML document with JavaScript, which dynamically loads the page content. While CSR reduces the server’s burden, it can lead to slower initial load times, as users must wait for JavaScript execution before meaningful content appears \cite{nextjs}.

>>>>

SSR generates HTML content on the server before sending it to the client,
resulting in faster initial page loads and better Search Engine Optimization
(SEO). However, SSR can increase server load and reduce
\textit{throughput}\footnote{The number of requests the server can handle per
  second (RPS)} since each request requires additional processing before
responding.

In contrast, CSR shifts the rendering workload to the browser. The server
initially sends a minimal HTML document with JavaScript, which dynamically
loads the page content. While CSR reduces the server’s burden, it can lead to
slower initial load times, as users must wait for JavaScript execution before
meaningful content appears.
